---
title: "aelab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{instruction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
rmarkdown.html_vignette.check_title = FALSE
library(aelab)
library(readxl)
library(tibble)
library(lubridate)
library(stats)
library(dplyr)
library(openxlsx)
library(ggplot2)
```

# Introduction

The goal of achieving net zero emissions by 2050 is a significant driver of research focused on understanding greenhouse gas (GHG) flux dynamics and the autotrophic production of ecosystems. This ambitious target highlights the urgent need to explore how different ecosystems contribute to GHG emissions and how they can be managed to reduce these emissions effectively. To support this research, this R package offers a variety of useful tools designed to facilitate the analysis of data related to GHG flux and production in aquatic environments.

# GHG flux calculation

Various techniques are used to measure greenhouse gas (GHG) flux, including eddy covariance (EC), the boundary layer method (BLM), and the static chamber method. This study provides R functions specifically designed to analyze data from GHG flux measurements using the static chamber method with a portable gas analyzer.
 
## Load and process the raw data file

First, we need to obtain necessary parameters from the raw data downloaded from the portable gas analyzer. We will remove unnecessary rows and columns and extract the GHG concentrations along with the corresponding date and time recorded by the analyzer using `tidy_ghg_analyzer()`. Additionally, this function will also eliminate any NaN values in the GHG data. Currently, this function supports raw data downloaded from:  
(i) LI-COR Trace Gas Analyzer (LI-7810 and LI-7820).
(ii) ABB LGR-ICOS Gas Analyzer (M-GGA-918).

```{r}
# The provided file is a raw data file downloaded from 
# the LI-COR Trace Gas Analyzer
ghg_data_path <- system.file("extdata", "ch4.xlsx", package = "aelab", mustWork = T)
ch4 <- tidy_ghg_analyzer(ghg_data_path, "ch4")
ch4[c(1:3), ]
```

Sometimes, the date and time recorded by the analyzer do not match the actual date and time we recorded in situ. Therefore, we need to convert the `date_time` column from the previous step to align with the real-life time, if there are any discrepancies:

```{r}
# The analyzer's time was assumed to be 
# 15 minutes and 30 seconds faster than real time
ch4 <- convert_time(ch4, min = -15, sec = 30)
ch4[c(1:3), c(5:6)]
```

## Calculate the slope

We first need to obtain the slope from the linear regression of GHG concentration changes over time within the chamber to calculate the flux. This can be done by defining the measurement time interval, either from a reference file (see Method 1 below) or directly in R (see Method 2 below).

### Method 1: Load from excel

In Excel, enter the date and time when the GHG flux measurement started, then load the file into R. Make sure the column containing the date and time information is formatted in ISO 8601 as `YYYY-MM-DD hh:mm:ss`.

```{r}
ref_data_path <- system.file("extdata", "reference.xlsx", package = "aelab", mustWork = T)
ref <- read_excel(ref_data_path)
ref
```

Now, we can perform regressions on the GHG concentrations using `calculate_regression()`. The reference file must contain a `date_time` column (measurement start times), a `site` column, and an `analyzer` column (e.g. `"1337"`).

### Window type

Two window modes are available via `window_type`:

- **`"sliding"`** (default): moves a fixed-length sub-window (default `num_rows = 300` rows ≈ 5 min) across the full measurement period and keeps the position with the highest R². Useful when the exact measurement start is uncertain.
- **`"fixed"`**: fits a single window of `duration_minutes` starting at `start_offset_s` seconds after the reference time. More reproducible — all gases are evaluated over the same interval. The first 30 s of chamber deployment are mechanically disturbed; skipping them with `start_offset_s = 30` is recommended.

```{r}
# Sliding window (default): best 5-min sub-window within 7-min period
calculate_regression(ch4, ref, ghg = "CH4", reference_time = "date_time",
                     site = "site", analyzer_code = "1337",
                     duration_minutes = 7, num_rows = 300)

# Fixed window: regress from t+30 s to t+7 min (skips dead-bar period)
calculate_regression(ch4, ref, ghg = "CH4", reference_time = "date_time",
                     site = "site", analyzer_code = "1337",
                     window_type = "fixed", start_offset_s = 30)
```

### Per-row window overrides

When `reference_df` contains a `duration_minutes` and/or `start_offset_s` column, a non-`NA` value in that column overrides the global parameter for that specific measurement. Rows without the column, or with `NA`, fall back to the global value. This is useful when individual closures have different deployment lengths or when one measurement needs a longer dead-bar skip.

### Model selection

The `fit_type` parameter controls the regression model:

- **`"linear"`** (default): ordinary least-squares slope. Valid when concentration change is small relative to ambient.
- **`"quadratic"`**: second-order polynomial; the reported slope is the initial rate at t = 0 (the coefficient of *t*).
- **`"exp_tz"`** / **`"exp_zhao18"`**: Hutchinson–Mosier nonlinear exponential models; slope evaluated at `t_zero`.
- **`"auto"`**: tries linear → quadratic → exp_tz in sequence, accepting a more complex model only when R² improves by > 0.01 and the exponential curvature parameter *b* does not exceed κ~max~ (Hüppi et al. 2018). The model chosen for each measurement is recorded in the `fit_used` output column.

When `fit_type = "auto"`, analyser precision is looked up automatically from the built-in `analyzer_precision` dataset (LI-7810, LI-7820, and LGR UGGA) if `instrument_precision` is not supplied. The lookup matches the `analyzer_code` argument against model names case-insensitively, so `"LGR"` correctly resolves to LGR UGGA precision rather than falling back to LI-7810.

```{r}
# Auto model selection with fixed window and 30-s dead-bar skip
calculate_regression(ch4, ref, ghg = "CH4", reference_time = "date_time",
                     site = "site", analyzer_code = "1337",
                     fit_type = "auto",
                     window_type = "fixed", start_offset_s = 30)
```

### Quality flags

Each measurement receives a `flag` in the output:

| Flag | Meaning |
|---|---|
| `ok` | R² ≥ threshold and correlation ≥ threshold; use this slope |
| `discard` | Detectable trend but poor fit (R² below threshold) |
| `zero` | No detectable trend (correlation below threshold, or slope within instrument noise) |
| `no_data` | Insufficient data in the window |

Two optional parameters tighten the QC:

- **`instrument_precision`**: analyser precision in ppm. Slopes whose absolute value falls below `precision / window_duration_s` are flagged `"zero"` even if R² passes. Also used by `fit_type = "auto"` to compute κ~max~ per measurement.
- **`g_factor_threshold`**: when `fit_type = "auto"` selects a nonlinear model, the g-factor (chosen slope / linear baseline slope) is computed and recorded in the `g_factor` output column. If `abs(g_factor) > g_factor_threshold`, the flag is overridden to `"discard"` (Gaudard et al. 2025). Default `NULL` (disabled).

### Method 2: Type directly in R

The reference information can also be constructed directly in R as a data frame.

```{r}
ref_direct <- data.frame(
  date_time = as.POSIXct("2023-03-11 07:32:00", tz = "Asia/Taipei"),
  site = "S1",
  analyzer = "1337"
)
calculate_regression(ch4, ref_direct, ghg = "CH4", reference_time = "date_time",
                     site = "site", analyzer_code = "1337")
```

### Method 3: Using `make_reference()`

`make_reference()` builds a `reference_df` from a compact multiline string — one closure per line, whitespace-separated: `time  site  analyzer  [dur=N]  [off=N]`. This avoids maintaining a separate Excel file for small campaigns or ad-hoc measurements.

```{r}
make_reference(
  date = "2023-03-11",
  measurements = "
    07:32:00  S1  1337
    07:45:00  S2  1337
  "
)
```

Per-row window overrides use `dur=` and `off=` tokens. Rows without an override simply omit the token.

```{r}
make_reference(
  date = "2023-03-11",
  measurements = "
    07:32:00  S1  1337  dur=3.5
    07:45:00  S2  1337  dur=2.0  off=45
    08:00:00  S3  1337
  "
)
```

When `output` is specified, the tibble is also written to `.xlsx`:

```{r eval=FALSE}
make_reference(
  date = "2023-03-11",
  measurements = "
    07:32:00  S1  1337
    07:45:00  S2  1337
  ",
  output = "reference.xlsx"
)
```

## Calculate the flux

The equation we used to calculate the GHG flux was derived from Yong et al. (2024):

$$
F = \frac{(S \times V \times c)}{R \times T \times A}
$$

where S is the slope obtained from the linear regression of GHG concentration changes over time (ppm s$^{-1}$), V is the chamber volume (liters), c is the conversion factor from seconds to hours, R is the ideal gas constant (0.082 L atm K$^{−1}$ mol$^{−1}$), T is the temperature inside the chamber (kelvin), and A is the surface area of the chamber (m$^2$).

Therefore, the function `calculate_ghg_flux()` requires a dataframe with columns for the slope, area, volume, and temperature. Note that the units of the parameters must match those described above, resulting in a flux with units of mmol m$^{-2}$ d$^{-1}$.

```{r}
results_ch4 <- calculate_regression(ch4, ref_direct, ghg = "CH4", reference_time = "date_time",
                                    site = "site", analyzer_code = "1337")
# In practice, filter to flag == "ok" before calculating flux
flux_ch4 <- data.frame(
    slope  = results_ch4$slope,
    area   = 1, # in square meter
    volume = 1, # in litre
    temp   = 1) # in celsius
calculate_ghg_flux(flux_ch4)
```

## Additional GHG tools

### `convert_ghg_unit()`

`convert_ghg_unit()` converts a GHG flux value to µg m⁻² h⁻¹. The input can be a plain number or a character string containing embedded numbers (e.g., a mean ± SD string); each number in the string is converted individually.

```{r}
# Numeric input: 97 mg CH4 m-2 h-1 → µg m-2 h-1
convert_ghg_unit(97, ghg = "ch4", mass = "mg", area = "m2", time = "hr")

# Character string input: each number is converted individually
convert_ghg_unit("0.002 ± 0.003", ghg = "ch4", mass = "mmol", area = "m2", time = "hr")
```

### `calculate_MDF()`

`calculate_MDF()` calculates the Minimum Detectable Flux (MDF) for a static chamber system. This is useful for evaluating whether a gas analyser is precise enough to detect fluxes at the expected magnitude of the study site.

```{r}
calculate_MDF(
  precision_ppm     = 1,
  closure_time_s    = 300,
  data_point_n      = 300,
  chamber_volume_m3 = 0.0064,
  temperature_C     = 25,
  chamber_area_m2   = 0.07
)
```

### `calculate_total_co2e()`

`calculate_total_co2e()` aggregates individual GHG fluxes (mg m⁻² h⁻¹) into a single total CO₂-equivalent flux (g m⁻² d⁻¹) using IPCC AR6 100-year GWPs (CH₄ = 27, N₂O = 273).

```{r}
calculate_total_co2e(co2 = 4.02, ch4 = 0.001, n2o = 0.003)
```

# NEM calculation

The change in oxygen concentration in the water reflects the balance between photosynthetic production, respiratory consumption, and physical exchange at the water-atmosphere interface. By measuring dissolved oxygen (DO) concentrations at high frequency with in situ data loggers, we can calculate gross primary production (GPP), ecosystem respiration (ER), and net ecosystem metabolism (NEM), as detailed by Staehr et al. (2010).

## Load and process the raw data file

First, we need to obtain the necessary parameters from the raw data downloaded from the data loggers. We will remove unnecessary rows and columns and extract the dissolved oxygen concentrations and water temperature, along with the corresponding date and time recorded by the logger using `process_hobo()`. This function will also eliminate any NA values in the dataframe, and for `type = "do"` it drops readings with a negative (sensor-error) DO or temperature value. It accepts both Chinese (上午/下午) and English (AM/PM or 24-hour) HOBOware locale exports. The `no_hobo` input was necessary for further analysis. `process_hobo()` supports two logger types via the `type` argument: `type = "do"` (default) for the HOBO U26 Dissolved Oxygen Data Logger, and `type = "temp"` for the HOBO Pro v2 Temperature/RH Logger (returns columns `air_temp` and `rh` instead of `do` and `temp`, with one row per reading).

For batch HOBO DO workflows, `combine_hobo()` reads a folder of logger exports, `filter_complete_days()` keeps only days with (near-)complete diel coverage, `add_sun_times()` attaches per-date sunrise/sunset from site coordinates (via suncalc), and `process_weather_mh()` / `combine_weather_mh()` import bulk hourly weather in the CWA CODiS "MH" format as an alternative to the per-day `process_weather()`.

```{r}
hobo_data_path <- system.file("extdata", "ex_hobo.csv", package = "aelab")
do <- process_hobo(hobo_data_path, no_hobo = "code_for_logger", type = "do")
do[c(1:3), ]
```

The air pressure and wind speed at the sampling site should be obtained either manually or from a nearby weather station. The `process_weather` function tidies the CSV file downloaded from Taiwan's weather station. In addition to extracting the columns with air pressure and wind speed values, it also duplicates the hourly data to match the half-hourly recorded frequency of the DO concentrations.

```{r}
weather_data_path <- system.file("extdata", "ex_weather.csv", package = "aelab")
weather <- process_weather(weather_data_path, date = "2024-04-10", zone = "zone_A")
weather[c(1:5), ]
```

In addition to the previously mentioned parameters, we also needed the water depth, salinity, start and end date and time of the data logger deployment, and the sunrise and sunset times during that period. These parameters can be entered in Excel and then imported into R using `process_info()` to ensure that all necessary parameters are included. The column `zone` corresponds to `process_weather()`, and column `no_hobo` corresponds to `process_hobo()`.

```{r}
info_data_path <- system.file("extdata", "info.xlsx", package = "aelab")
info <- process_info(info_data_path)
info
```

Finally, we can merge the dataframe containing air pressure and wind speed data from `process_weather()` with the processed data downloaded from the data loggers. Next, we will combine this with the other necessary parameters using the `zone` and `no_hobo` columns. We can use the `plot_hobo()` function to inspect the changes in DO concentrations.

```{r}
data <- merge(weather, do, by = "date_time")
merged_df <- data %>% 
  inner_join(info, by = c("zone", "no_hobo"))
merged_df[c(1:3), ]
plot_hobo(merged_df)
```

## Calculate GPP and ER

GPP and ER can be calculated using `calculate_do()`. NEM (denoted as `nep` in the output dataframe) was calculated by GPP + ER.

```{r}
calculate_do(merged_df)
```

# Statistical analysis

aelab provides a small suite of statistical helpers that cover a typical aquatic ecology workflow: describe your data, check for outliers, test normality, transform if needed, and choose between a parametric or non-parametric group comparison.

The examples below use a simple self-contained data frame with three groups.

```{r}
set.seed(42)
stat_df <- data.frame(
  group = rep(c("A", "B", "C"), each = 8),
  value = c(rnorm(8, 2, 0.4), rnorm(8, 3.5, 0.5), rnorm(8, 5, 0.6))
)
```

## Descriptive statistics — `descriptive_statistic()`

`descriptive_statistic()` returns mean ± SD and min–max for each group.

```{r}
descriptive_statistic(stat_df, vars = value, groups = group)
```

## Outlier detection — `find_outlier()`

`find_outlier()` uses the IQR method (values outside 1.5 × IQR from Q1/Q3) to flag outliers.

```{r}
find_outlier(stat_df, var = "value")
```

## Normality testing

Use `normality_test_t()` when comparing two groups (t-test context), and `normality_test_aov()` when comparing three or more groups (ANOVA context). Both apply Shapiro-Wilk to raw, square-root, and log10 transforms so you can identify whether a transformation improves normality.

```{r}
# Two-group context
normality_test_t(stat_df, "value", group, "A", "B")

# One-way ANOVA context (all three groups)
normality_test_aov(stat_df, "value", "group")
```

## Data transformation — `df_trans()`

If residuals are not normal, `df_trans()` appends a transformed column to the data frame for re-testing.

```{r}
stat_df_t <- df_trans(stat_df, "value", "sqrt")
head(stat_df_t)
```

## Parametric test — `aov_test()`

When normality holds, `aov_test()` runs a one-way ANOVA followed by Tukey HSD post-hoc comparisons and returns compact letter display (CLD) groupings.

```{r}
aov_test(stat_df, "value", "group")
```

## Non-parametric test — `ks_test()`

When normality cannot be assumed, `ks_test()` runs Kruskal-Wallis followed by Dunn's post-hoc test with Bonferroni correction.

```{r}
ks_test(stat_df, "value", "group")
```

## Faceted CLD annotation — `sig_labels()`

`sig_labels()` wraps `ks_test()` to produce compact letter display (CLD) annotations for faceted ggplot2 boxplots. It iterates over each level of a faceting variable (`by`, default `"year"`), runs `ks_test()` on each subset, and returns a tidy data frame ready to pass directly to `geom_text()`. Facets where the Kruskal-Wallis p-value is ≥ `alpha` are silently dropped.

```{r results='hide'}
set.seed(1)
sig_df <- data.frame(
  year = rep(c("2023", "2024"), each = 20),
  site = rep(c("A", "B", "C", "D"), 10),
  flux = c(
    rnorm(20, mean = rep(c(1, 3, 2, 6), 5)),
    rnorm(20)
  )
)
labels <- sig_labels(sig_df, "flux", "site", by = "year")
```

```{r}
labels
```

Only 2023 appears in `labels` because 2024 has a non-significant Kruskal-Wallis result. Pass the output to `geom_text()` via the `data` argument to annotate a faceted boxplot:

```{r}
ggplot(sig_df, aes(x = site, y = flux)) +
  geom_boxplot() +
  geom_text(
    data   = labels,
    aes(x = site, y = y_pos, label = Letter),
    vjust  = -0.5
  ) +
  facet_wrap(~year)
```

When the raw observations used for testing differ from the plotted data (e.g. the plot shows aggregated values), pass the plotted data as `plot_data` so y-positions are computed on the correct scale.

# Visualization

aelab provides plot wrappers — `plot_point()`, `plot_line()`, `plot_box()`, and `plot_bar()` — with a consistent theme built on Century Gothic. Custom colour palettes are available through `aelab_palettes()`, `scale_fill_aelab_d()`, and related scale functions.

The example below previews the `"ghg"` palette and applies it to a box plot comparing CH₄ fluxes across three habitat types.

```{r eval=FALSE}
# Preview the "ghg" palette
aelab_palettes("ghg")

# Box plot with aelab palette
df_vis <- data.frame(
  site = rep(c("Mangrove", "Mudflat", "Seagrass"), each = 10),
  ch4  = c(rnorm(10, 2, 0.5), rnorm(10, 1, 0.3), rnorm(10, 0.5, 0.2))
)

plot_box(df_vis, site, ch4, site) +
  scale_fill_aelab_d("ghg")
```

# Reference

Gaudard, J., Telford, R. J., & Halbritter, A. H. (2025). fluxible: An R package to process ecosystem gas fluxes from closed-loop chambers in an automated and reproducible way. Methods in Ecology and Evolution, 00, 1–9. https://doi.org/10.1111/2041-210X.70161

Hüppi, R., Felber, R., Krauss, M., Six, J., Leifeld, J., & Fuß, R. (2018). Restricting the nonlinearity parameter in soil greenhouse gas flux calculations for more reliable flux estimates. PLOS ONE, 13(7), e0200876. https://doi.org/10.1371/journal.pone.0200876

Staehr, P. A., Bade, D., Van de Bogert, M. C., Koch, G. R., Williamson, C., Hanson, P., Cole, J. J., & Kratz, T. (2010). Lake metabolism and the diel oxygen technique: State of the science. Limnology and Oceanography: Methods, 8(11), 628–644. https://doi.org/10.4319/lom.2010.8.0628

Yong, Z.-J., Lin, W.-J., Lin, C.-W., & Lin, H.-J. (2024). Tidal influence on carbon dioxide and methane fluxes from tree stems and soils in mangrove forests. Biogeosciences, 21(22), 5247–5260. https://doi.org/10.5194/bg-21-5247-2024
